---
title: "Draft 2"
author: "Ana Karina"
date: "4/7/2020"
#output: html_document
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

# Risk of infection from touching surfaces contaminated with SARS-CoV-2
## Abstract
## Introduction
## Materials and Methods 
### Hazard Identification
Whenever an infected individual coughs, sneezes or speaks, droplets containing pathogens make their way to the proximal surfaces [@Bourouiba2014]. It has been shown that SARS-CoV-2 can persist on surfaces like plastic, metal, or glass at room temperature for hours to days [@VanDoremalen2020]. Additionally, viruses are known to transfer from surfaces to the hands upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991], as well as from hands to the lips and saliva of individuals [@Rusin2002; @Pitol2017]. Given the fact that people frequently touch their faces and the mucous membranes within the face (nose, eyes and mouth) [@Kwok2015; @Nicas2008], we have identified the indirect (surface-mediated) transfer of SARS-CoV-2 as a relevant transmission pathway. Therefore, this risk assessment is designed to explore the risks of infection associated with the surface-mediated transmission of SARS-CoV-2. 

### Exposure Assessment 
#### Scenario definition
Only one potential exposure pathway was analysed: the indirect transmission of viruses through contaminated surfaces. Two different scenarios where modelled: single touch (frequently touched buttons) and multiple touches (ATM or cash machine). The stochastic and mechanistic model used to model the risk of surface-mediated transmission of SARS-CoV-2 infection is based on models described elsewhere [@Julian2018; @Julian2019]. 

*Scenario 1. Single touch buttons.* Frequently touched buttons such as traffic light buttons, train buttons, and elevator buttons are of risk concern due to their frequent use by multiple people. In this scenario, we modelled the risk of touching a contaminated button followed by touching mucous membranes on the face. The material of the button was assumed to be either plastic or steel. The sequence of events leading to the viral inoculation of the susceptible individual was considered to be the following: 1) contamination event (infected person coughing directly on the surface), 2) virus decay on the surface, 3) virus transfer from the surface to the finger, 4) virus transfer from the finger to the mucous membranes  of the nose, mouth and eyes.

*Scenario 2. Multiple touches.* ATM/cash machines are another example of surfaces that are frequently visited by individuals and are not frequently cleaned, as they are located in public spaces. For this risk assessment model, the ATM was assumed to be made of steel. The sequence of events leading to virus inoculation was considered to be the following: 1) contamination event (infected person coughing directly on the surface), 2) virus decay on the surface, 3) virus transfer from the surface to the finger and from finger to surface through multiple contacts, 4) virus transfer from the finger to the mucous membranes in the face (eyes, mouth, nose).


####  Virus concentration and transfer 
Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input. The virus titers used in the risk assessment were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020]. The RT-qPCR method estimates the number of copies of viral RNA present in the samples. Nevertheless, the presence of coronavirus RNA does not imply the presence of infectious viruses. In risk assessments whenever data on the concentration of infectious viruses is not available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses [@Chigor2014]. The gene copy number to infectivity ratio is, to our knowledge, not available for SARS-CoV-2. In the absence of SARS-CoV-2 data, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016 (Table 1). Both, influenza viruses and coronaviruses, are enveloped, RNA viruses with similar sizes. 

The concentration of viruses in the surface after inoculation (through cough) was estimated using the following equation: 

\begin{equation}
C(0)=\frac { C_{sp}V_{s}}{\frac {4 \pi x_{i}^{2}}{a}}
\end{equation}

where $C(0)$ [virus/cm^2^] is the concentration of virus in the surface, $V_{s}$ [mL/cough] is the volume of saliva expelled per cough, $C_{sp}$ [virus/mL] is the concentration of infective viruses in the saliva or sputum, $x_i$ [cm] is the inoculation distance or the distance between the surface and the infected individual at the time of the cough for activity $i$, and $a$ is the XXX. This equation assumes that the droplets are expelled in a conical shape and a proportion of the spheric part of the cone is in contact with the surface. After surface inoculation, the concentration of the virus on the surface at time = $t$ was estimated using the following equation:

\begin{equation}
C(t)= C(0) e^{-nt}
\end{equation}

where $C(t)$ [virus/cm^2^] is the concentration of infective viruses at time $t$, $C(0)$ [virus/cm^2^] is the initial concentration of virus in the surface, $t$[min] is the time lapsed after surface inoculation, and $n$ [m^-1^] is the decay rate of the virus in the selected surface (metal or plastic). The concentration of virus on the hand after a single touch was estimated to be a function of the transfer efficiency of viruses between the hands and the surfaces. The concentration of viruses in the hand after multiple touches was estimated as reported elsewhere [@Julian2009; @Julian2018], where the transfer of virus is a function of the gradient in the concentration of viruses bwteen the hand and the surface. 

\begin{equation}
C_{f_f} = C(t) TE_{sh}
\end{equation}

or 

\begin{equation}
C_{f_f} = (C(t)-C_{f_i}) TE_{sh} + C_{f_i}
\end{equation}

where $C_{f_i}$ and $C_{f_f}$ [virus/cm^2^] are the initial and final concentration of virus in the finger, $C(t)$ [virus/cm^2^] is the concentration of virus in the surface at time = $t$, $TE_{sh}$[%] is the transfer efficiency of viruses between surfaces and hands. The dose of viruses that entered the susceptible individual through facial membranes, $D$ [virus], was estimated as follows: 

\begin{equation}
D = C_{f_f} S_{fm} TE_{hm}
\end{equation}

where $C_{finger_f}$ [virus/cm^2^] is the final concentration of virus in the finger, $S_{f}$ [cm] is the surface area of the finger in contact with the mucous membranes, and $TE_{hm}$ [%] is the transfer efficiency of the virus from the hand to the mucous membranes. 

#### Dose-response
In the case of highly virulent pathogens, where no prophylaxis is available, it is common practice to use data of animal trials to generate dose-response models. The dose-response relationship used in this model was obtained from the QMRA wiki <http://qmrawiki.org/experiments/sars> and is based on two studies [@DeAlbuquerque2006; @DeDiego2008] of SARS-CoV and Murine hepatitis virus (MHV-1) infection in mice. MHV is a commonly used surrogate for human coronaviruses. The risk of infection was estimated with the following equation:

\begin{equation}
P_{inf}=1 - e^{-kD}
\end{equation}

where $P_{inf}$ [%] is the probability of infection, $k$ [-] is the infectivity parameter of the exponential model and $D$ [virus] is the dose of virus that entered the individual through mucous membranes. 


#### Model parameters used in the risk assessment 
| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
|:--:|:--:|:--:|:--:|:--:|:--:|
| $C_{spGC}$ | gene copies/mL | Concentration of SARS-CoV-2 in sputum or saliva | Log~10~ gene copies/mL (data set) | ReSample | [@Wolfel2020; @Pan2020; @Kim2020; @To2020] RT-qPCR data from 9, 2, 2, and 23 patients with COVID-19|
| $GC:TCID_{50}$ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | [@Ip2016] Mean RT-qPCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
| $C_{sp}$ | TCID~50~/mL |Concentration of infective SARS-CoV-2 in sample | $C_{sp}=\frac {C_{spGC}}{GC:TCID_{50}}$ | - | calculated|
| $V_s$ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | [@Nicas2009] volume = 0.044mL, [@Adhikari2019] uniform $\pm$ 10%|
| $N_{cgh/min}$ |min^-1^ | Number of coughs per minute  | 0.57(1)| Normal |[@Leung2020], based on data of 17 individuals with seasonal CoV| 
|$t_{ATM}$ | min |Time spent on the ATM | 0.75(0.38) | Normal | This study. Otained from 20 hours of observational data |
|$t_{btw-ATM}$ | min |Time between visits | 0.13, 0.80 | Weibull | This study. Otained from 20 hours of observational data |
|$t_{btw-btn}$ | min |Time between pressing button | 1-20 | Uniform | Assumed |
|$x_{ATM}$| cm | ATM inoculation distance | 40 | Point value | This study. Assumed from the observational data  |
|$x_{btn}$| cm | Button inoculation distance | 50 | Point value | Assumed |
|$t_{\frac {1}{2}stl}$| min | Half life of CoV-2 in steel | 338(35) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
|$t_{\frac {1}{2}pl}$| min | Half life of CoV-2 in plastic | 409(39) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
| $n_x$| min^-1^ |Decay rate in surface $x$ (steel or plastic) | $n_x=\frac {ln(2)}{t_\frac{1}{2}}$ | - | calculated |
| $TE_{sh-stl}$ | % | Transfer efficiency of virus between steel to hand | 37.4(16) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
| $TE_{sh-pl}$ | % | Transfer efficiency of virus between plastic to hand | 79.5(21.2) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
| $TE_{hm}$ | % | Transfer efficiency from hand to mucouse membranes | 20(6.3) | Normal | [@Pitol2017], transfer efficiency of viruses (MS2) from hand to saliva | 
| $A_{f}$ | cm^2^ | Finger surface area | 3.9-5.9 | Uniform | Fractional surface area of partial finger [@AuYeung2008] times average hand surface area [@EPA2011]| 
| $k$ | PFU^-1^ | Parameter of the exponential dose-response| 0.00107, 0.00135, 0.00680 | Triangular | Data obtained from QMRAwiki, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 0.5th, 50th, and 99.5th percentiles as min, mode, and max |
|:--:|:--:|:--:|:--:|:--:|:--:|
The input parameters of the distributions are: Normal: mean (SD); Uniform: lower-bound - upper-bound; Triangular: min, mode, max; Weibull: scale, shape.

### Risk Characterization 
Monte Carlo simulation: method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.

### Sensitivity analysis

A sensitivity analysis was performed to investigate hot the variabilty and uncertainty of the parameters in the model onfluenced the estimated risks. The sensitivity was estimated using Spearman correlation coefficients between the inputs and outputs of the model.

## Results and Discussion
```{r dose-response, include=FALSE}

```

## Supporting Information

## Author Information

## Aknowledgements 

## References 


