---
title: "Draft 1"
author: "Ana Karina"
date: "4/7/2020"
output: html_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

# Risk of infection from touching surfaces contaminated with SARS-CoV-2
## Abstract
## Introduction
close contact route (large droplet), airborne (small droplets), fomites.
An investigation of a nosocomial outbreak of COVID-19 reported at Augustine’s Hospital in Durban, South Africa suggest that indirect contact through health care workers and fomites were the predominant transmission pathways @Lessells2020.

*Risk assessment models*
Mathematical models have been used to understand SARS-CoV and MERS-CoV @Adhikari2019 @Lei2017.
Let et al @Lei2017  developed a model to understand the relative contribution of different transmission routes (airborne, close contact, and fomites) on the risk of infection for influenza A H1N1, SARS-CoV, and norovirus. The simulation results suggest that around 50% of the CoV infections were due to indirect (fomite mediated) transmission. 
A QMRA of MERS-CoV was developed by Adhikari et all @Adhikari2019 to evaluate the risk for hospital transmission of coronavirus. They only considered the risk of infection due from droplets and aerosolized particles, since that is the main transmission route. Most of the uncertainty associated with the model was due to the concentration of virus in the saliva. The researchers evaluated possible interventions and showed that the use of mask contributed significantly to the reduction of the risk (>90%), as well as increasing air exchange (up to 58% risk reduction).

extensive environmental contamination has been found in settings with COVID-19 patients @Ong2020 @Ye2020, coronaviruses persist on inanimate surface for extended periods of time @Kampf2020

If no countermeasures are taken, viruses can spread through the space populated by the human host at a high speed, as we have observed @BarOn2020.

Surfaces play a significant role in the spread of respiratory viral diseases @Boone2007

Potential role of aerosol transmission of SARS-CoV-2, sampling the air @Liu2020

The CDC and WHO have highlighted two main transmission routes: direct transmission or person-to-person transmission (respiratory droplets produced while talking, coughing, or sneezing), and indirect transmission or surface-mediated transmission (touching a contaminated surface followed by touching mucose membranes in the face such as mouth or nose) @WHO2020 @CDC2020. 

Although indirect transmission through fomite mediated contacts is not the primary transmission route, it can lead to infection @Chen2004. A study of an outbreak of SARS CoV in Taiwan showed that, out of the 232 SARS-CoV patients in the hospital, 31 patients (13 %)  were hospital-acquired transmissions. Out of the 31 cases, most of the transmission occurred by direct contact with a known SARS patient, but 6/31 (~20%) of the transmissions could not be traced to direct contact with SARS patients. Sampling of surfaces for the RNA of coronavirus showed that surfaces such as drinking water buttons, beds and chairs were contaminated with coronavirus. Thus, the researchers postulated that indirect contact could have been the source of transmission @Chen2004.

Viruses are known to transfer from contaminated surfaces to the skin upon contact @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991 and from the skin to the mouth and saliva of individuals @Pitol2017 @Rusin2002. People touch their faces frequently @Nicas2008. Therefore, touching contaminated surfaces offers a risk of exposure. Despite its importance, there is not a lot of information regarding the magnitude of risk associated with touching contaminated surfaces @BarOn2020.

Hand hygiene interventions and surface disinfection have been highlighted as one of the primary strategies to control transmission @Mcclellan2020. Studies with coronaviruses have shown than hygiene practices such as handwashing can reduce the risk of coronavirus infection @Beale2020. But there is no conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.



## Materials and Methods 
### Hazard Identification
When an individual infected with SARS-CoV-2 coughs, sneezes, or speaks, there could be particles of saliva containing the virus that make their way to the proximal surfaces. SARS-CoV-2 can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020. Viruses are known to transfer from surfaces to the skin upon contact  @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991. It is also well known that people frequently touch their faces and the mucouse membranes within the face (nose, eyes and mouth) @Kwok2015. Therefore, indirect transmission of SARS-CoV-2 a relevant transmission pathway. Despite its relevance, there are no studies quantifying the risks associated with indirect transmission of SARS-CoV-2 between surfaces and people. Hence, this risk assessment is designed to explore the risks associated indirect (fomite mediated) transmission of SARS-CoV-2. Since the probability of illness due to infection is, up to date, not known (REF), the end point of this risk assessment is probability of infection. 

### Exposure Assessment 
Two expossure scenarios were modeled. 
*Scenario 1. Single touch.*
Frequently touched buttons (Traffic light buttons, train buttons). High touched surfaces and Supermarket products. 

*Scenatio 2. Multiple touches.*
ATM machines.

(Insert diagram)

#### Parameters table
| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
|:--:|:--:|:--:|:--:|:--:|:--:|
| C~sp~ | Gene Copies/ml | Concentration of SARS-CoV-2 in sputum or saliva | log~10~ [4,4.8...] | ReSample | @Wolfel2020 @Pan2020 @Kim2020 @To2020, RT-PCR data from 9, 2, 2, and 23 patients with COVID-19|
| GC:TCID~50~ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | @Ip2016, mean RT-PCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
| V~slv/cgh~ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | @Nicas2009, volume = 0.044, @Adhikari2019, assumed uniform $\pm$ 10%|
| N~cgh/min~ |min^-1^ | Number of caughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV| 
|t~ATM~ | min |Time spend doing the activity *i* | [...] | Normal | This study. Activity ATM: 20 hours of observational data |
|x~ATM~| m | Inoculation distance for activity *i* | [...] | Point value |  This study. Estimated based on activity ATM: 20 hours of observational data |
| k | PFU^-1^ | Parameter of the exponential dose-response| 0.00246, 0.00135, 0.00459 | Triangular | Data obtained from QMRA wiki http://qmrawiki.org/experiments/sars, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 5th, 50th, and 95th percentiles as min, mode, and max |
| n~stl~ | min^-1^ |Decay rate in steel | [...] | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
| n~pl~ | min^-1^ |Decay rate in plastic | [...] | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
| TE~sh_stl~ | % | Transfer efficiency from surface (steel) to hand | Normal | 37.4 (16) | @Lopez2013, transfer efficiency of viruses (MS2) from surfaces to skin at 40-65% humidity | 
| TE~sh_pl~ | % | Transfer efficiency from surface (plastic) to hand | Normal | 79.5 (21.2) | @Lopez2013, transfer efficiency of viruses (MS2) from acrylic to skin at 40-65% humidity | 
| TE~hm~ | - | Transfer efficiency from hand to mucouse membrane | Normal | 0.201(0.063) | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
| A~sf~ | cm^2^ | Finger surface area |Uniform | 3.8-6.6 | Data from @Chabrelie2018 @ @Peters2009a @Sahmel2015 | 
| FSA~fm~ | - | Fractional surface area | Uniform | 0.5-0.8 | Assumed | 


#### SARS-CoV-2 Concentration in saliva samples
There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals. Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data. We used data on the viral loads of the saliva or sputum of an infected person within the first 14 days of symptom onset @Wolfel2020 @Pan2020 @Kim2020 @To2020.

Althoug translating copy numbers into infectivity data is not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF). 

In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016. Both, influenza viruses and coronaviruses are enveloped, RNA viruses with similar sizes. 

*Conversion factor RT-PCR data to infectivity*
Although the ratio of infectivity to gene copy numbers is, to our knowledge, available for coronavirus, there is information of this ratio for the samples taken from patients for other respiratory viruses such as different strains of influenza virus @Yan2018 @Ip2016.

GC:Infectivity
@Yan2018 significant correlation between virus RNA copies and infectious virus recovered from breathe and swabs of influenza patients,

#### SARS-CoV-2 Concentration in surfaces

"viru transferred between two surfaces is often formulated as the product of contact frequency, transfer efficiency, surface concentration, and contact area"

\begin{equation}
C(0)=\frac {V_{slv/cgh} C_{sp}}{4 \pi x_{i}^{2}  0.2}
\end{equation}


\begin{equation}
C(t)= C(0)  e^{-nt}
\end{equation}

#### Dose Response for SARS-CoV

To "To estimate the infection risk based on the dose received,"
“For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”

The dose-response relationship was obtain from QMRA wiki <http://qmrawiki.org/experiments/sars>. The model is based on animal dose-response relationship. No human dose-response relationship is available for SARS Coronavirus. It is generally accepted to use animal models in QMRA if human data is not available @Haas2014, @Haas2015.

\begin{equation}
P_{inf}=1 - e^{-kd}
\end{equation}

### Risk Characterization 
Monte Carlo simulation method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.

### Sensitivity analysis

A sensitivity analysis was performed to investigate hot the variabilty and uncertainty of the parameters in the model onfluenced the estimated risks. The sensitivity was estimated using Spearman correlation coefficients between the inputs and outputs of the model.


( Insert table of parameters )

## Results and Discussion
```{r dose-response, include=FALSE}

```

Since no randomized control trials will be runned and there are limits to the conclusions we can reach using epidemiological data, risk assesments provide an attempt to understand the risk associated with different activities. The results of this paper could inform policy. For example, trafick lights in the crossing stations could clausured momentaniely, the same than train buttons.

The dependence of temperature and humidity in the transfer rate of coronavirus is unclear. If temperature and humidity may play a role in non-dominant transmission pathways, such as indirect transfer, it would be hard to observe epidemiological studies. It is known that the survival of viruses on surfaces is dependent on environmental factors such as temperature and humidity (REF). Similarly, the transfer rate of viruses varies with humidity (REF). Nevertheless, if only a small fraction of the transmission occurs through indirect transfer we wont see it. 

The sensitivity analysis shows that XXX parapeters are responsible for XX % of the risk variation of the model. Collecting more data on these parameters could reduce uncertainty.

It is important to notice that the effectiveness of interventions such as hand washing depends not only on many factors which include: the hand-wasihng technique, the suceptibility of the virus to phisical and chemical desinfection, and the relative contribution of the indirect transfer of viruses to the overal infection risk. As it has been pointed out, thand hygiene should be promoted but the risk of overemphasizing it without more intormation could turn out to be negative (Yang, 2020). ...  ("risk compensation"?).

Seasonal factors such as temperature, humidity and sunlight are known to influence the spread of respiratory viruses such as influenza (REF). The role that those factos have on the spread of SARS-CoV-2 remains unknown (REF), althouh it is know that the temperature and humidity influences the viability of the virus (REF), as well as SARS-CoV-2 survive longer at lower temperature (REF).

It has been shown that seasonal environmental factors such as temperature and humidity influence respiratory virus survival and transmission @Moriyama2020.

Humidity influences the size of the droplets generated by the infected individual, the size of the droplets influence both the lifetime of the virus in the environment. 
The deposition of droplets onto surfaces depends on the particle size.

Nevertheless, our results suggest that fomite mediated transmission will be .. influenced by those factors. The extend that this influence will have in the droplet dynanimcs by influencing evaporation rate of the droplets (REF). transmission of CoV will depend on the relative contribution of indirect transmission pathway. 

### Limitations

Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020. The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF). In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes. 

Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters. 

We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countribution of each route. 

Even though this risk assessment model relies on many assumptions, some with high uncertainty around them, and the risk caucluation is hihgly simplidied, this model can: provide insigts into the possible effects that different interventions can have in reducing the spread of SARS-CoV-2, serve as a first step in the pathway of building more complex models that allow exploring the relative contribution that different transmission routes could have in spreddin the disease. 


Currently we have liminted data on the relationship between copy numbers and infectivity, which limits the extend to which the risk assessment can be used. However, modeling approaches such as the QMRA can help elucidate the contribution of different exposure routes to the disease transmission and therefore help the development of guidelines. 

_____
about the paper, we can say something like "Although the relative importance of each transmission route remains poorly understood, there has been a growing consensus that contaminated fomites play a critical role in the spread of viruses." @Boone2007

The relative contribution of surface-mediated transmission as compared with other transmission routes such as direct transmission, air, or thorugh the feca-oral route remains to be addressed. QMRA can be used to incorporate different transmission routes and answer this question.



Although not perfect, a risk assessment framework could be use to conduct a hazzard assessment. Since offices will need to re-open 
https://www.osha.gov/shpguidelines/hazard-Identification.html.

A more comprehensive model such as the models described elsewhere (REF), could potentially help elucidating the relative contribution of different expossure routes, and inform the population regading the most sensitive measures they could take given the particularities of the situation (e.g., air conditionated in offices)


## Supporting Information

## Author Information

## Aknowledgements 

## References 


